Predicting Conflict Escalation

### 2023-04-30 docs/hes_final/article.qmd
Formatting Fixes and new spike fig

@@

To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512xn$ matrix for each country and month where $n$ is the number of articles published in that month related to that country. Each of these matrices was then split into several random subsamples of 50 articles each. Samples for countries in which $n<50$ were padded to $512x50$. This process had two effects on training. First, by limiting the size of each sample to 50 articles, we reduced the dimensionality of each sample. Simultaneously, by splitting the corpus into multiple unique training examples per country per month, we were able to substantially increase the size of the training dataset.

@@

With the encoder's training complete, we returned to the original base dataset. Each country-month matrix was again split into $512x50$ matrices. Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector. Each vector was then stacked, resulting in a single training example for each country and month of size $16x\frac{n}{50}$. This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target. 
In contrast to our statistical modeling approach, the transformer models are trained purely on text embeddings. The model was not conditioned by country, and no additional features were added beyond the text corpus itself.

@@

One of the potential research extensions of this work is to continue experimenting
with data sampling and aggregation approaches. We used either monthly averages 
to create $512x1$ vectors or article level samples to create $512x50$ matrices
in our dataset preparation. While we received some promising results, it is possible 
that the first approach makes the data too coarse and the second too noisy. Another
alternative could be generating daily averages and combining them into $512x31$ (largest number of days in a month)
matrices. This could potentially be more sensitive to the signal than the monthly averages
and less noisy than article level samples.

>>>>

![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png)

@@

To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512\times n$ matrix for each country and month where $n$ is the number of articles published in that month related to that country. Each of these matrices was then split into several random subsamples of 50 articles each. Samples for countries in which $n<50$ were padded to $512\times50$. This process had two effects on training. First, by limiting the size of each sample to 50 articles, we reduced the dimensionality of each sample. Simultaneously, by splitting the corpus into multiple unique training examples per country per month, we were able to substantially increase the size of the training dataset.

@@

With the encoder's training complete, we returned to the original base dataset. Each country-month matrix was again split into $512\times50$ matrices. Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector. Each vector was then stacked, resulting in a single training example for each country and month of size $16\times\frac{n}{50}$. This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target.
In contrast to our statistical modeling approach, the transformer models are trained purely on text embeddings. The model was not conditioned by country, and no additional features were added beyond the text corpus itself.

@@

One of the potential research extensions of this work is to continue experimenting
with data sampling and aggregation approaches. We used either monthly averages 
to create $512\times1$ vectors or article level samples to create $512\times50$ matrices
in our dataset preparation. While we received some promising results, it is possible 
that the first approach makes the data too coarse and the second too noisy. Another
alternative could be generating daily averages and combining them into $512\times31$ (largest number of days in a month)
matrices. This could potentially be more sensitive to the signal than the monthly averages
and less noisy than article level samples.
